---
title: "DNA length in protein-DNA complexes"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

**Last updated on `r date()`.**

## How to re-use this work

If you use these figures in your own work, please cite this website: <https://doi.org/10.5281/zenodo.3470119>

## Structures of protein-DNA complexes by experimental method

All figures are interactive (you can zoom in, and hovering over elements will
show more information).

```{r Download and clean up data, generate summary}
# Load required packages
library(magrittr)
library(jsonlite)
library(dplyr)
library(forcats)
library(stringr)
library(ggplot2)
library(plotly)
library(here)

# Query the PDB for all DNA molecules in crystal structures of protein-DNA
# complexes
pdb_query <- 'https://www.ebi.ac.uk/pdbe/search/pdb/select?q=molecule_type:%22DNA%22%20AND%20assembly_composition:%22DNA/protein%20complex%22&fl=pdb_id,molecule_sequence,experimental_method&rows=1000000&wt=json'

pdb_data <- pdb_query %>% 
    fromJSON() %>% 
    .$response %>% 
    .$docs %>% 
    as_tibble() %>% 
    distinct(pdb_id, .keep_all = TRUE) %>% 
    filter(experimental_method %in% c("X-ray diffraction",
                                      "Electron Microscopy",
                                      "Solution NMR")) %>% 
    mutate(dna_length = str_length(molecule_sequence),
           experimental_method = as_factor(as.character(experimental_method)))

# Summary statistics
protein_dna_cplx_structures <- ggplot(data = pdb_data) +
    geom_bar(mapping = aes(x = experimental_method)) +
    theme_bw() +
    xlab("") +
    ylab("Number of PDB entries") +
    ggtitle("Structures of protein-DNA complexes")
ggplotly(protein_dna_cplx_structures)
```

```{r Save SVG file methods, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "protein-dna-cplx-structures.svg",
       plot = protein_dna_cplx_structures,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/protein-dna-cplx-structures.svg)

## DNA length in crystal structures of protein-DNA complexes
### Entire distribution

```{r DNA length in crystal structures}
dna_length_xtal <- pdb_data %>% 
    filter(experimental_method == "X-ray diffraction") %>% 
    ggplot() +
    geom_histogram(mapping = aes(x = dna_length), binwidth = 1) +
    theme_bw() +
    ggtitle("DNA length in crystal structures of protein-DNA complexes") +
    xlab("DNA length (bp)") +
    ylab("Number of crystal structures")
ggplotly(dna_length_xtal)
```

```{r Save SVG file xtal, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "dna-length-in-protein-dna-cplx-xtal-structures.svg",
       plot = dna_length_xtal,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/dna-length-in-protein-dna-cplx-xtal-structures.svg)

### 0-150 bp range

```{r DNA length in crystal structures in 0-150 bp range}
dna_length_xtal_150 <- pdb_data %>% 
    filter(experimental_method == "X-ray diffraction" & dna_length < 151) %>% 
    ggplot() +
    geom_histogram(mapping = aes(x = dna_length), binwidth = 1) +
    theme_bw() +
    ggtitle("DNA length in crystal structures of protein-DNA complexes (0-150 bp)") +
    xlab("DNA length (bp)") +
    ylab("Number of crystal structures")
ggplotly(dna_length_xtal_150)
```

```{r Save SVG file xtal 0-150 bp range, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "dna-length-in-protein-dna-cplx-xtal-structures-0-150.svg",
       plot = dna_length_xtal_150,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/dna-length-in-protein-dna-cplx-xtal-structures-0-150.svg)

## DNA length in cryo-EM structures of protein-DNA complexes

```{r DNA length in cryo-EM structures}
dna_length_cryoem <- pdb_data %>% 
    filter(experimental_method == "Electron Microscopy") %>% 
    ggplot() +
    geom_histogram(mapping = aes(x = dna_length), binwidth = 1) +
    theme_bw() +
    ggtitle("DNA length in cryo-EM structures of protein-DNA complexes") +
    xlab("DNA length (bp)") +
    ylab("Number of cryo-EM structures")
ggplotly(dna_length_cryoem)
```

```{r Save SVG file cryoem, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "dna-length-in-protein-dna-cplx-cryoem-structures.svg",
       plot = dna_length_cryoem,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/dna-length-in-protein-dna-cplx-cryoem-structures.svg)

## DNA length in NMR structures of protein-DNA complexes

```{r DNA length in NMR structures}
dna_length_nmr <- pdb_data %>% 
    filter(experimental_method == "Solution NMR") %>% 
    ggplot() +
    geom_histogram(mapping = aes(x = dna_length), binwidth = 1) +
    theme_bw() +
    ggtitle("DNA length in NMR structures of protein-DNA complexes") +
    xlab("DNA length (bp)") +
    ylab("Number of NMR structures")
ggplotly(dna_length_nmr)
```

```{r Save SVG file nmr, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "dna-length-in-protein-dna-cplx-nmr-structures.svg",
       plot = dna_length_nmr,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/dna-length-in-protein-dna-cplx-nmr-structures.svg)

## Dataset

The histograms presented above are derived from the following dataset:

```{r Entire dataset}
# Format table for display
pdb_table <- pdb_data %>% 
    arrange(desc(dna_length)) %>% 
    select(`PDB code` = pdb_id,
           `DNA length` = dna_length)
pdb_table
```


```{r Save dataset as JSON, include=FALSE}
# Save dataset for download
if (!dir.exists(here("datasets"))) {
    dir.create(here("datasets"))
}

write_json(pdb_data, here("datasets", "protein-dna-cplx-structures.json"))
```

[**Download raw dataset in JSON format**](datasets/protein-dna-cplx-structures.json)
