---
title: "SARS-CoV-2 structures"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

**Last updated on `r date()`.**

## How to re-use this work

If you use these figures in your own work, please cite this website:
<https://doi.org/10.5281/zenodo.3470119>

```{r Load packages, download data and prepare dataset}
# Load required packages
library(magrittr)
library(dplyr)
library(jsonlite)
library(forcats)
library(ggplot2)
library(plotly)
library(here)
library(lubridate)

# We need two queries: one for uppercase titles, the other for lowercase ones
pdb_query <- 'https://www.ebi.ac.uk/pdbe/search/pdb/select?q=title:*SARS-CoV-2*%20AND%20status:REL&fl=pdb_id,deposition_date,citation_year,title,experimental_method,resolution,organism_scientific_name,molecule_name,molecule_type&rows=1000000&wt=json'

# This is a helper pipeline to extract data
dig_up_data <- . %>%
    .$response %>%
    .$docs %>%
    as_tibble()

# Query the PDB and clean up data
pdb_data <- pdb_query %>% 
    fromJSON() %>% 
    dig_up_data() %>% 
    bind_rows() %>% 
    mutate(
      experimental_method   = as_factor(as.character(experimental_method)),
      deposition_date       = as_date(ymd_hms(deposition_date)),
      citation_year         = as.integer(citation_year),
      molecule_name         = as.character(molecule_name),
      molecule_type         = as_factor(as.character(molecule_type))
      ) %>% 
    distinct(pdb_id, .keep_all = TRUE)
```

## About this section

With the ongoing [COVID-19
pandemic](https://en.wikipedia.org/wiki/COVID-19_pandemic), new structures of
proteins from
[SARS-CoV-2](https://en.wikipedia.org/wiki/Severe_acute_respiratory_syndrome_coronavirus_2) are deposited to the PDB at a fast pace. The visualizations presented here
won't help curb the pandemic (and are not medical advice, obviously), but I hope
they will make curious and helpless minds (like myself) get a feel for the
research effort currently being deployed against it.

The following resources are a lot more relevant and actionable:

- [Coronavirus Structural Task Force](https://insidecorona.net/)
- [PDBe COVID-19 Data Portal](https://www.ebi.ac.uk/pdbe/covid-19)
- [Crowdfight COVID-19](https://crowdfightcovid19.org/)

## SARS-CoV-2 structures by deposition date

All figures are interactive (you can zoom in, and hovering over elements will
show more information).

```{r SARS-CoV-2 structures by year}
sars_cov2_structures_year <- pdb_data %>% 
    ggplot() +
    geom_bar(mapping = aes(x = deposition_date)) +
    ggtitle("Structures of SARS-CoV-2 by deposition date") +
    xlab("Deposition date") +
    ylab("Number of PDB entries") +
    theme_bw()
ggplotly(sars_cov2_structures_year)
```

The two noticeable spikes (68 structures deposited on March 15th and 13
structures deposited on April 7th) come from [a fragment screening
experiment](https://www.diamond.ac.uk/covid-19/for-scientists/Main-protease-structure-and-XChem.html)
performed at Diamond Light Source. All these structures are the same protein
(main protease) in complex with candidate molecules for drug design.

```{r Save SVG file of year graph, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "sars-cov2-structures-by-year.svg",
       plot = sars_cov2_structures_year,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/sars-cov2-structures-by-year.svg)

## SARS-CoV-2 structures by experimental method

```{r SARS-CoV-2 structures by experimental method}
sars_cov2_structures_method <- pdb_data %>% 
    ggplot() +
    geom_bar(mapping = aes(x = experimental_method, fill = experimental_method)) +
    guides(fill = guide_legend(title = "Experimental method")) +
    ggtitle("Structures of SARS-CoV-2 by experimental method") +
    xlab("") +
    ylab("Number of PDB entries") +
    theme_bw()
ggplotly(sars_cov2_structures_method)
```

```{r Save SVG file of method graph, include=FALSE}
# Save figure for download
if (!dir.exists(here("figures"))) {
    dir.create(here("figures"))
}
ggsave(filename = "sars-cov2-structures-by-method.svg",
       plot = sars_cov2_structures_method,
       device = "svg",
       path = here("figures"))
```

[**Download figure in SVG format**](figures/sars-cov2-structures-by-method.svg)


## Dataset

The graphs presented above are derived from the following dataset:

```{r Entire dataset}
# Format table for display
pdb_table <- pdb_data %>% 
    arrange(desc(citation_year)) %>% 
    select(`PDB code` = pdb_id,
           `Citation year` = citation_year,
           `Experimental method` = experimental_method,
           Title = title)
pdb_table
```

```{r Save dataset as JSON, include=FALSE}
# Save dataset for download
if (!dir.exists(here("datasets"))) {
    dir.create(here("datasets"))
}

write_json(pdb_data, here("datasets", "sars-cov2-structures.json"))
```

[**Download raw dataset in JSON format**](datasets/sars-cov2-structures.json)
